---
title: "Fake European Data"
output: github_document
---

```{r}
library("here")
library("readr")
library("dplyr")
library("ggplot2")
library("lubridate")
```

The purpose of this document is to manufacture a 15-minute time-series that spans two-years, 2015-2016, in the context of the `"Europe/Paris"` timezone. By using such a time series, I will have a way to do calclulations in Vega, while checking them in R.

```{r}
tz_local <- "Europe/Paris"

t_mean <- 10.0
dt_day <- 7.5
dt_year <- 10.0

dtm_ref <- parse_datetime("2015-01-01 00:00:00", locale = locale(tz = tz_local))
per_day <- ddays(1)
per_year <- ddays(365.25)
```

To give us a little structure, our fake time-series will have daily variation and yearly variation. Let's imagine that this is a temperature, in 째C (this *is* Europe). Our mean temperature will be `r t_mean` 째C, with a daily variation (above and below the daily mean) of `r dt_day` 째C, and a yearly variation of `dt_year` 째C. The coldest intstant of the year is at midnight January 1st. The daily period is 24 hours; the yearly period is 365.25 (24-hour) days.

```{r}
n_obs <- ((dtm_ref + years(2)) - dtm_ref) / dminutes(15)

fake_temperature_paris <- 
  tibble(
    instant = dtm_ref + seq(0, n_obs - 1)*dminutes(15),
    temperature = 
      t_mean - 
      dt_day * cos(2 * pi * (instant - dtm_ref)/per_day) - 
      dt_year * cos(2 * pi * (instant - dtm_ref)/per_year)  
  )
```

```{r}
make_plot <- function(df) {
  ggplot(df, aes_string(x = "instant", y = "temperature")) +
  geom_line()  
}

fake_temperature_paris %>%
  make_plot()
```

```{r}
fake_temperature_paris %>%
  filter(year(instant) == 2015, month(instant) == 1) %>%
  make_plot()
```

So this looks about right...

Let's write our data out to a csv file:

```{r}
write_csv(
  fake_temperature_paris, 
  here("docs", "data", "fake_temperature_paris.csv")
)
```
