---
title: "Timezones: Vega, Vega-Lite, Python and R"
output: github_document
---


Until I can wrap my head around all the concepts, this is going to be me "puking up" all of my knowledge and opinions on timezones. It is hoped that this can, with enough consideration and revision, serve as a foundation for efforts to improve the time-handling capabilities of Vega and Vega-Lite.

We want a way to render a spec independent of the browser-locale.

```{r}
library("magrittr")
library("readr")
library("glue")
```

## Time

The first thing I want to sort out is how time is handled in R and Python. I think this is a useful prelimanary exercise because chances are that a Vega/Vega-Lite spec is giong to be built using R or Python.

I think it can also be useful so we can understand, a bit better, each other languages.

### R

In R, we have a type `POSIXct`, which is timezone-aware. Internally, it stores the number of seconds since the UNIX epoch. There is an attribute, `tzone`, that may or may not be set (in my opinion, it should be set).

First, I'm going to make a helper function to look a little at the internals of a given `POSIXct`.

```{r}
describe <- function(x) {
  
  as_utc <- x
  attr(as_utc, "tzone") <- "UTC"
  
  print(
    glue::glue("UTC time:       {as_utc}"),
    glue::glue("Formatted time: {x}"),
    glue::glue("Timezone:       {attr(x, 'tzone')}"),
    glue::glue("Numeric value:  {as.numeric(x)}")
  )
  
  invisible(x)
}
```

---

Let's start with parsing a string into a POSIXct, and not specifying a timezone:

```{r}
"2013-04-05 06:07:08" %>%
  as.POSIXct() %>%
  describe()
```

It looks like the system parses this as a local time (I am running this from the `"America/Chicago"` timezone), so it is stored as the UTC time that corresponds to local time described by the string.

---

Next, let's specify the timezone to the constructor:

```{r}
"2013-04-05 06:07:08" %>%
  as.POSIXct(tz = "UTC") %>%
  describe()
```

This makes sense, and is repeatable, not depending on my system settings.

---

Let's give the constructor an ISO-8601 string:

```{r}
"2013-04-05T06:07:08Z" %>%
  as.POSIXct() %>%
  describe()
```

This is not good. Not good at all. It is using the system time and has choked on the `"T"` in the string.

---

Luckily we have an alternative, the `readr::parse_datetime()` function.

```{r}
"2013-04-05T06:07:08Z" %>%
  readr::parse_datetime() %>%
  describe()
```

Here, we recognize the ISO-8601 string, and it sets the timezone to `"UTC"`. Using Pandas' vocabulary, we have localized to `"UTC"`.

---

When parsing, we can set a timezone:

```{r}
"2013-04-05 06:07:08" %>%
  readr::parse_datetime(locale = readr::locale(tz = "America/Chicago")) %>%
  describe()
```

If the format is ambiguous (no `"Z"` or `"+00:00"` at the end), the parser localizes to the timezone provided.

---

What if the string is already implies the localization to be UTC?

```{r}
"2013-04-05T06:07:08Z" %>%
  readr::parse_datetime(locale = readr::locale(tz = "America/Chicago")) %>%
  describe()
```

In this case, the parser localized to `"UTC"`, then (again using Pandas' vocabulary) converts to `"America/Chicago"`.

---

When writing a spec directly from R, it becomes the R users' responsibility to make sure that POSIXct is serialized to an ISO format, and to serialize the timezone somehow (whenever that may become meaningful).

### Python

Here's where I need to educate myself a little bit, so I'll follow Jake's lead:

https://github.com/vega/vega-lite/issues/4044#issuecomment-408565320

> One thing that would be worth thinking about is whether we can support `datetime64[ns,utc]` as well as `datetime64[ns]` in a streamlined way. That way people could choose UTC or not within Python, and everything else should just flow from there.

I'll start with a Python version of my `describe()` function.

```{python}
# Python
import pandas as pd

def describe(x):

    if x.tzinfo is None:
      utc_time = 'not defined'
    else:
      utc_time = x.tz_convert('UTC')
    
    print('UTC time:       ', utc_time)
    print('Formatted time: ', x)
    print('Timezone:       ', x.tzinfo)
    print('Numeric value:  ', x.value)
    print('')
    print(pd.DataFrame({'x' : [x]}).dtypes)
    return x
```

---

First, let's try an ambiguous format:

```{python}
x = pd.to_datetime('2013-04-05 06:07:08')

describe(x)
```

This brings a number of things into focus - my apologies for noting things that should be evident from reading the documentation.

The first is that what is being stored is the number of nanoseconds from an epoch. Because we did not fix a timezone, it seems we are using a "local" timezone where this is the number of seconds from an epoch defined in that timezone.

```{python}
x = pd.to_datetime('2013-04-05T06:07:08Z').tz_localize('UTC').tz_convert('America/Chicago')

describe(x)
```

```{python}
x = pd.to_datetime('2013-04-05T06:07:08Z')

describe(x)
```

### reticulate

```{r}
library("reticulate")

pd <- import("pandas")
```

Let's try sending data to Python, testing if the timezone gets sent over:

```{r}
df <- 
  data.frame(
    x = readr::parse_datetime(
      "2013-04-05T06:07:08Z", 
      locale = readr::locale(tz = "America/Chicago")
    )
  )

describe(df$x)

pd_df <- r_to_py(df)

cat("\n")
pd_df
pd_df$dtypes
```

Let's set the timezone there and bring it back

```{r}
pd_df_new <- pd_df$copy()
pd_df_new$x <- pd_df_new$x$dt$tz_localize("UTC")
pd_df_new$x

pd_df_new$x <- pd_df_new$x$dt$tz_convert("America/New_York")
pd_df_new$x

df_new <- py_to_r(pd_df_new)
describe(df_new$x)
```

## Intake

Jake VanderPlas:

> I wouldn't want to make Pandas output UTC because then, for example, if I make a chart in Seattle and send it to a friend in NYC the rendering will be different.

For me, the takeaway is that we want a consistent rendering regardless of where a chart is rendered, and to support and document how to do just that.


Would like more detail on how the [data format](https://vega.github.io/vega-lite/docs/data.html#format) works in Vega/Vega-Lite. For example, the documention implies that JavaScript's `Date.parse()` is used in [Vega-Lite](https://vega.github.io/vega-lite/docs/data.html#format), but not in [Vega](https://vega.github.io/vega/docs/data/#format).

Look to https://github.com/altair-viz/altair/pull/1053 for how Jake uses timezones. 

Also, look to https://github.com/altair-viz/altair/pull/1053#issuecomment-408610839 for the problem of browser location. (Earlier Jake comment)

It can be useful to have a slightly different interpretation of date vs. datetime.

The timezone should be an attribute of the data itself. It should be available at the top level: see https://github.com/vega/vega-lite/issues/4004

Vega can parse to `"boolean"`, `"date"`, `"number"` or `"string"`. So, in Vega, is a datetime just a number that represents the number of milliseconds from the UNIX epoch? If we are to associate a timezone with a datetime in the parsing, it would seem to require its own type. Maybe this one of reasons behind [this comment](https://github.com/vega/vega-lite/issues/4044#issuecomment-406023278). 

In efforts to decouple the data from the visualization specification, it becomes our responsibility to provide both the serialized dataframe (csv, json, etc.) and the parsing specification (timezones).
 
