---
output: github_document
---

Could `r_to_py()` and `py_to_r()` be adapted such that is the source data has columns with timezone attributes, that the timezone is set in the target data?

Could this be done without stepping on #160? (tagging @DavisVaughn) On this question, I wonder if a distinction can be made between tz-naive `datetime64[ns]` and tz-aware `datetime64[ns, <tz>]`.

---

My experience with Python is limited, so maybe the best thing I can do is to demonstrate. In my Rmd, I am using both R and Python chunks, so I will label those explicitly to appear in the Markdown.

```{r}
# R
library("reticulate")
library("readr")
# imports "glue"

pd <- import("pandas")

# function to describe the attributes of a POSIXct:
describe <- function(x) {
  
  tz <- attr(x, "tzone")
  
  if (identical(tz, "")) {
    as_utc <- "not defined"
  } else {
    as_utc <- x
    attr(as_utc, "tzone") <- "UTC"    
  }

  str_value <- formatC(as.numeric(x) * 1e3, digits = 13, width = 13)
  
  print(
    glue::glue("UTC time:       {as_utc}"),
    glue::glue("Formatted time: {x}"),
    glue::glue("Timezone:       {tz}"),
    glue::glue("Milliseconds:   {str_value}")
  )
  
  invisible(x)
}
```

```{python}
# Python
import pandas as pd

def describe(x):

    if x.tzinfo is None:
      utc_time = 'not defined'
    else:
      utc_time = x.tz_convert('UTC')
    
    print('UTC time:       ', utc_time)
    print('Formatted time: ', x)
    print('Timezone:       ', x.tzinfo)
    print('Milliseconds:   ', x.value / 1.e6)
    print('')
    print(pd.DataFrame({'x' : [x]}).dtypes)
    return x
```

---

For the first step, let's create a data frame in R that we will send to Python. It will have a single column with a single value. 

```{r}
# R
df_r_to_py <- 
  data.frame(
    x = parse_datetime(
      "2013-04-05T06:00:00Z", 
      locale = locale(tz = "America/Chicago")
    )
  )

describe(df_r_to_py$x)
```

So far, so good.

---

Let's see what Python sees:

```{python}
# Python
describe(r.df_r_to_py.x[0])
```

We see that the numerical representation has been "preserved", but that the localization has not been brought over. 

---

In Python, let's create a copy of what was sent from R, then let's do the localization manually. 

```{python}
# Python
df_py_to_r = r.df_r_to_py.copy()
df_py_to_r.x = df_py_to_r.x.dt.tz_localize('UTC').dt.tz_convert('America/Chicago')

describe(df_py_to_r.x[0])
```

This looks good now.

---

Finally, let's look at this in R:

```{r}
# R
df_return <- py$df_py_to_r

describe(df_return$x)
```

Here, there seems to be partially good-news. The numerical value has been preserved, and the timzone is set.

The "bad" news is that the timezone is set to `"UTC"`, but this is one of the behaviors sought in #160 (but perhaps only for the tz-naive case). 

## Summary

Just to restate from the top:

Could `r_to_py()` and `py_to_r()` be adapted such that is the source data has columns with timezone attributes, that the timezone is set in the target data?

Could this be done without stepping on #160? On this question, I wonder if a distinction can be made between tz-naive `datetime64[ns]` and tz-aware `datetime64[ns, <tz>]`.

--------

```{r}
# R
df_r_to_py_naive <- data.frame(x = as.POSIXct("2013-04-05 06:00:00"))


describe(df_r_to_py_naive$x)
```

```{python}
# Python
describe(r.df_r_to_py_naive.x[0])
```

```{python}
# Python
df_py_to_r_naive = r.df_r_to_py_naive.copy()
```

```{r}
describe(py$df_py_to_r_naive$x)
```